---
title: "Reproducible Research - Week 2 Course Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='figure/')
```

## Loading and preprocessing the data

Load the data and process/transform the data into a format suitable for your analysis.

```{r}
# load the data
activity <- read.csv("./data/activity.csv", header = TRUE)

# use head, str and summary functions to get a better understanding of the data
head(activity)
str(activity)
summary(activity)

# date column is of type Factor, convert it to date format and check that conversion worked with str function
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
str(activity)
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day

```{r}
# calculate steps taken per day, ignoring missing values
activity_nona <- na.omit(activity)
steps_per_day <- tapply(activity_nona$steps, activity_nona$date, FUN=sum, na.rm=TRUE)
head(steps_per_day)
```

2. Make a histogram of the total number of steps taken each day

```{r}
# using base to create histogram
hist(steps_per_day, xlab = "Total Steps Taken Per Day", ylab = "Frequency", main = "Histogram of Total Steps Taken Per Day", breaks=15)
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean(steps_per_day)
median(steps_per_day)
```

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
library(dplyr)
average_steps <- activity %>% select(interval, steps) %>% na.omit() %>% group_by(interval) %>% summarize(avgsteps= mean(steps)) 
plot(average_steps$interval, average_steps$avgsteps, xlab="5-minute interval", ylab="Average number of steps taken", main="Average steps taken per 5-minute interval", type="l")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
average_steps[which.max(average_steps$avgsteps), ]
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset 

```{r}
activity_na <- is.na(activity$steps)
sum(activity_na)
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc

```{r}
# Fill in missing values with the mean number of steps for that interval (excluding NAs)
replace_data <- function(x) {
  replace(x, is.na(x), mean(x, na.rm = TRUE))
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
activity_2 <- activity %>% group_by(interval) %>% mutate(steps = replace_data(steps))
head(activity_2)

# check that there are no NAs left in data
activity_na_2 <- is.na(activity_2$steps)
sum(activity_na_2)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
steps_per_day_2 <- tapply(activity_2$steps, activity_2$date, FUN=sum, na.rm=TRUE)
hist(steps_per_day_2, xlab = "Total Steps Taken Per Day", ylab = "Frequency", main = "Histogram of Total Steps Taken Per Day", breaks=15)
mean(steps_per_day_2)
median(steps_per_day_2)
```

The mean of this new dataset is the same as the mean calculated in the first part of the assignment because in the original data, the NAs were removed and were not part of the mean calculation, whereas now we have added additional rows to the data where the steps are based on the average of that same dataset, resulting in the exact same mean as before.

However, the median is now higher because we have added additional rows using the mean of the original data, pushing up the median to be exactly the same as the mean in the new dataset.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r}
activity_2$day <- weekdays(activity_2$date)
activity_2$weekday_or_weekend <- ifelse(activity_2$day %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

```{r}
average_steps_2 <- activity_2 %>% select(interval, steps, weekday_or_weekend) %>% group_by(interval, weekday_or_weekend) %>% summarize(avgsteps= mean(steps))

library(ggplot2)
ggplot(average_steps_2, aes(x = interval, y = avgsteps)) + 
    geom_line() +
    facet_grid(.~ weekday_or_weekend) + 
    xlab("5-minute interval") + 
    ylab("Average number of steps taken") +
    ggtitle("Average number of steps taken on Weekdays vs Weekends")
```

